---
title: "Human and Economic impact of a selection of weather events in the US (1950 - 2011)"
output: html_document
---

# Synopsis
Storms and other severe weather events can cause both public health and economic problems for communities and municipalities. Many severe events can result in fatalities, injuries, and property damage, and preventing such outcomes to the extent possible is a key concern. This project involves exploring the U.S. National Oceanic and Atmospheric Administration's (NOAA) storm database. This database tracks characteristics of major storms and weather events in the United States, including when and where they occur, as well as estimates of any fatalities, injuries, and property damage.

# Data processing 

Load the packages that will be used throughout the study

```{r, echo = TRUE}
library(plyr)
library(ggplot2)
```

Load the data from the Coursera website; download it into your working directory, unzip the file, and import the dataset into R. 

```{r, echo = TRUE}
path <- getwd()
url <- "https://d396qusza40orc.cloudfront.net/repdata%2Fdata%2FStormData.csv.bz2"
download.file(url, file.path(path, "repdata-data-StormData.csv.bz2"))
Storm_Data <- read.csv(bzfile("repdata-data-StormData.csv.bz2"))
```

Take a first glance at the data

```{r, echo = TRUE}
names(Storm_Data)
```

# Results

Find the deadliest weather events 

```{r}
## Find the deadliest weather event
fatalityData <- aggregate(FATALITIES ~ EVTYPE, data = Storm_Data, FUN ="sum")
most_fatal <- which.max(fatalityData$FATALITIES)

fatalityData[most_fatal,]

## Find the 10 deadliest events
fatalityData <- arrange(fatalityData, desc(fatalityData[,2]))
top10fatData <- fatalityData[1:10,]
top10fatData
```

Plot the 10 deadliest weather events 

```{r, echo = TRUE}
ggplot(top10fatData, aes(x = reorder(EVTYPE, -FATALITIES), y = FATALITIES)) + 
  geom_bar(stat = "identity") + 
  ggtitle("10 deadliest events") + 
  xlab("Event Type") +
  theme(axis.text.x = element_text(angle = 45, hjust =1))
```


Find the weather events that lead to the most injured people

```{r, echo = TRUE}
## Find the event that leads to the most injured people
injuriesData <- aggregate(INJURIES ~ EVTYPE, data=Storm_Data, FUN = "sum")
most_injuries <- which.max(injuriesData$INJURIES)

injuriesData[most_injuries,]

## Find the 10 events that lead to the most injured people
injuriesData <- arrange(injuriesData, desc(injuriesData[,2]))
top10injData <- injuriesData[1:10,]
top10injData
```

Plot the 10 events that lead to the most injured people

```{r, echo = TRUE}
ggplot(top10injData, aes(x = reorder(EVTYPE, -INJURIES), y = INJURIES)) + 
  geom_bar(stat = "identity") + 
  ggtitle("10 events leading to the most injuries") + 
  xlab("Event Type") +
  theme(axis.text.x = element_text(angle = 45, hjust =1))
```


Find the weather events that have the biggest negative impact on the economy (cost the most). 
Impact on the economy is characterized in this study by PROPDMG (the property damages) and CROPDMG (crop damages).

```{r, echo = TRUE}
EconomicImpactData <- aggregate(PROPDMG + CROPDMG ~ EVTYPE, data = Storm_Data, FUN=sum)
names(EconomicImpactData) <- c("EVTYPE", "Damages")

EcoImpact <- arrange(EconomicImpactData, desc(EconomicImpactData[, 2]))
top10EcoData <- EcoImpact[1:10,]
top10EcoData
```

Plot the 10 weather events the have the biggest negative impact on the economy (cost the most). 

```{r, echo = TRUE}
ggplot(top10EcoData, aes(x = reorder(EVTYPE, -Damages), y = Damages)) + 
  geom_bar(stat = "identity") + 
  ggtitle("10 events having the greatest economic damages") + 
  xlab("Event Type") +
  theme(axis.text.x = element_text(angle = 90))
```

